---
title: "Project Part 1"
author: "EB VERSION"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(Sleuth3)
library(leaps)
library(stats)
library(car)
library(corrplot)
library(dplyr)
library(ggplot2)
library(GGally)
library(gridExtra)
library(psych)
library(naniar)
library(VIM)
library(mice)
library(DMwR2)
library(knitr)


```



## Data Set
```{r water potability}
#attach(water_potability)

water_potability <- read.csv("water_potability.csv", header = TRUE)

head(water_potability)

str(water_potability)
#NOTE: Keep potability integer or factor?

water_potability$Potability <- factor(water_potability$Potability,
  levels = c(0, 1),
  labels = c("Non-Potable", "Potable")
)

describe(water_potability)  
# NOTE: Z-score these to better compare?  or unneccessary?

vis_miss(water_potability)

matrixplot(water_potability, 
           main = "Water Potability Missing Data Matrix")


na_percentage <- water_potability %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100))
print(na_percentage)
# Most NA values are ph, sulfate, and Trihalomethanes


```

Potability is factored with 2 levels: "Non-Potable" and "Potable"

4.4% of values are missing, mostly in Sulfates and ph levels with a smaller portion in Trihalomehanes.


```{r correlation}


water_potability_noNA = na.omit(water_potability)
dim(water_potability_noNA)  # 2011 observations, 10 variables
dim(water_potability)  # 3276 observations, 10 variables

# omitting NA values reduced data set from 3276 observations to 2011 observations
# NOTE: consider removing what observations are necessary

vis_miss(water_potability_noNA)
#confirmed no missing values remain, however we have lost over 1000 observations

r <- water_potability_noNA %>% dplyr::select(where(is.numeric)) %>% cor()
corrplot(r)

# Solid set of variables with little to no collinearity

summary_of_missing <- miss_var_summary(water_potability)
print(summary_of_missing)

naniar::vis_miss(water_potability)
VIM::aggr(water_potability)

par(mar = c(10, 6, 3, 3))
aggr(water_potability, cex.axis = 0.6, numbers = TRUE, sortVars = TRUE)

par(mar = c(30, 10, 2, 2))
md.pattern(water_potability, rotate.names = TRUE)

gg_miss_upset(water_potability) 


# Create imputation for missing values using KNN
water_potability_knn <- knnImputation(water_potability, k = 5)

# Fit LDA with imputed data
lda_model <- lda(Potability ~ ., data = water_potability_knn)
print(lda_model)

lda_coeffs <- as.data.frame(lda_model$scaling)
lda_means <- as.data.frame(lda_model$means)

kable(lda_coeffs, caption = "LDA Coefficients")
kable(lda_means, caption = "LDA means")

# Fit QDA with imputed data
qda_model <- qda(Potability ~ ., data = water_potability_knn)
print(qda_model)

qda_coeffs <- as.data.frame(qda_model$scaling)
kable(qda_coeffs, caption = "QDA Coefficients")


#Examine density plots for imputed vs non-imputed data distribution comparisons.

ggplot() + geom_density(data = water_potability, aes(x = ph), fill = "red", alpha = 0.5) + geom_density(data = water_potability_knn, aes(x = ph), fill = "blue", alpha = 0.5) + 
  labs(title = "Distribution of pH values prior to (red) and after (blue) KNN imputation",
       x = "pH",
       y = "Density") +
  theme_minimal()

ggplot() + geom_density(data = water_potability, aes(x = Sulfate), fill = "red", alpha = 0.5) + geom_density(data = water_potability_knn, aes(x = Sulfate), fill = "blue", alpha = 0.5) + 
  labs(title = "Distribution of Sulfate values prior to (red) and after (blue) KNN imputation",
       x = "pH",
       y = "Sulfate") +
  theme_minimal()

ggplot() + geom_density(data = water_potability, aes(x = Trihalomethanes), fill = "red", alpha = 0.5) + geom_density(data = water_potability_knn, aes(x = Trihalomethanes), fill = "blue", alpha = 0.5) + 
  labs(title = "Distribution of Trihalomethanes values prior to (red) and after (blue) KNN imputation",
       x = "pH",
       y = "Trihalomethanes") +
  theme_minimal()

ggplot() + geom_density(data = water_potability, aes(x = Solids), fill = "red", alpha = 0.5) + geom_density(data = water_potability_knn, aes(x = Solids), fill = "blue", alpha = 0.5) + 
  labs(title = "Distribution of Solids values prior to (red) and after (blue) KNN imputation",
       x = "pH",
       y = "Solids") +
  theme_minimal()



```

No dark blue or red associations. 

A correlation greater than .6 is considered a strong correlation, and greater than .8 a strong relationship. 

Also, no NA values remain in the dataset, however we have lost over 1000 observations


```{r models}

head(water_potability)

# Everything Model
model1 = glm(Potability ~ ph + Hardness + Solids + Chloramines + Sulfate + Conductivity + Organic_carbon + Trihalomethanes + Turbidity, data = water_potability, family = binomial(link = "logit"), na.action = na.omit)
summary(model1) # Nothing is significant

# chemical Model
model2 = glm(Potability ~ Chloramines + Sulfate + Trihalomethanes, data = water_potability, family = binomial(link = "logit"), na.action = na.omit)
summary(model2) # Nothing is significant

# Simplest Model
model3 = glm(Potability ~ Organic_carbon, data = water_potability, family = binomial(link = "logit"), na.action = na.omit)
summary(model3) # Nothing is significant, solids close, organic_carbon kinda, 


```


Graphs are important for two reasons. They allow us to assess non-linear relationships, and also categorical variables.

```{r exploration}

water_potability %>% dplyr::select(where(is.numeric)) %>% pairs()

water_potability %>% 
  select(Potability, names(water_potability)[2:9] ) %>% 
  ggpairs()

```



```{r categoricalvariable}

head(water_potability)

#PLOTS for full dataset (no NA removal)

plot1 <- ggplot(water_potability, aes(x = Potability, y = ph, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by ph",
       x = "Potability",
       y = "pH") +
  theme_minimal() +
  theme(legend.position = "none")

plot2 <- ggplot(water_potability, aes(x = Potability, y = Hardness, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Hardness",
       x = "Potability",
       y = "Hardness") +
  theme_minimal() +
  theme(legend.position = "none")

plot3 <- ggplot(water_potability, aes(x = Potability, y = Solids, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Solids",
       x = "Potability",
       y = "Solids") +
  theme_minimal()  +
  theme(legend.position = "none")

plot4 <- ggplot(water_potability, aes(x = Potability, y = Chloramines, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Chloramines",
       x = "Potability",
       y = "Chloramines") +
  theme_minimal() +
  theme(legend.position = "none")

plot5 <- ggplot(water_potability, aes(x = Potability, y = Sulfate, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Sulfate",
       x = "Potability",
       y = "Sulfate") +
  theme_minimal() +
  theme(legend.position = "none")

plot6 <- ggplot(water_potability, aes(x = Potability, y = Conductivity, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Conductivity",
       x = "Potability",
       y = "Conductivity") +
  theme_minimal() +
  theme(legend.position = "none")

plot7 <- ggplot(water_potability, aes(x = Potability, y = Organic_carbon, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Organic_carbon",
       x = "Potability",
       y = "Organic_carbon") +
  theme_minimal() +
  theme(legend.position = "none")

plot8 <- ggplot(water_potability, aes(x = Potability, y = Trihalomethanes, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Trihalomethanes",
       x = "Potability",
       y = "Trihalomethanes") +
  theme_minimal() +
  theme(legend.position = "none")

plot9 <- ggplot(water_potability, aes(x = Potability, y = Turbidity, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Turbidity",
       x = "Potability",
       y = "Turbidity") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(
  plot1, plot2, plot3,
  plot4, plot5, plot6,
  plot7, plot8, plot9,
  ncol = 3
)

#PLOTS for dataset with NA removal

plot1 <- ggplot(water_potability_noNA, aes(x = Potability, y = ph, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by ph",
       x = "Potability",
       y = "pH") +
  theme_minimal() +
  theme(legend.position = "none")

plot2 <- ggplot(water_potability_noNA, aes(x = Potability, y = Hardness, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Hardness",
       x = "Potability",
       y = "Hardness") +
  theme_minimal() +
  theme(legend.position = "none")

plot3 <- ggplot(water_potability_noNA, aes(x = Potability, y = Solids, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Solids",
       x = "Potability",
       y = "Solids") +
  theme_minimal()  +
  theme(legend.position = "none")

plot4 <- ggplot(water_potability_noNA, aes(x = Potability, y = Chloramines, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Chloramines",
       x = "Potability",
       y = "Chloramines") +
  theme_minimal() +
  theme(legend.position = "none")

plot5 <- ggplot(water_potability_noNA, aes(x = Potability, y = Sulfate, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Sulfate",
       x = "Potability",
       y = "Sulfate") +
  theme_minimal() +
  theme(legend.position = "none")

plot6 <- ggplot(water_potability_noNA, aes(x = Potability, y = Conductivity, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Conductivity",
       x = "Potability",
       y = "Conductivity") +
  theme_minimal() +
  theme(legend.position = "none")

plot7 <- ggplot(water_potability_noNA, aes(x = Potability, y = Organic_carbon, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Organic_carbon",
       x = "Potability",
       y = "Organic_carbon") +
  theme_minimal() +
  theme(legend.position = "none")

plot8 <- ggplot(water_potability_noNA, aes(x = Potability, y = Trihalomethanes, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Trihalomethanes",
       x = "Potability",
       y = "Trihalomethanes") +
  theme_minimal() +
  theme(legend.position = "none")

plot9 <- ggplot(water_potability_noNA, aes(x = Potability, y = Turbidity, fill = Potability)) +
  geom_boxplot() +
  labs(title = "Potability by Turbidity",
       x = "Potability",
       y = "Turbidity") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(
  plot1, plot2, plot3,
  plot4, plot5, plot6,
  plot7, plot8, plot9,
  ncol = 3
)

ggplot(water_potability, aes(x = Solids, color = Potability)) +
  geom_density() +
  labs(title = "Density Plot of Solids & Potability",
       x = "Solids",
       y = "Density") +
  theme_minimal()




# Compute variance proportions from singular values
lda_variance <- data.frame(
  LD = paste0("LD", 1:length(lda_model$svd)),
  Proportion_of_Variance = lda_model$svd^2 / sum(lda_model$svd^2)
)

# Display as a table
kable(lda_variance, caption = "Proportion of Variance Explained by Each Linear Discriminant")


```




