---
title: "Final Project: Water Potability"
Author: "Antonia Sunseri and Evrim Baykal"
output: html_document
date: "2025-01-22"
---

Data set: https://www.kaggle.com/datasets/adityakadiwal/water-potability.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS) # For LDA and QDA analysis
library(Sleuth3)
library(car)
library(corrplot)
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyverse)
library(caret) # For train-test split and model evaluation
library(pROC)  # For ROC curve analysis
```

##Project Beginning: 
*State the data set you will be studying and the goal of the study. You will need to specify the output variable of interest as well as your main analytical method you intend to use.*

  The data set that we will analyze contains water quality metrics for 3,276 different bodies of water. The information contained within this data set will be used to address the concern for access to safe drinking-water. Access to safe drinking water is essential for health and a component of effective policy for health protection. Thus, analyzing water quality metrics is important as a health and development issue at a national, regional, and local level. The output variable of interest will be water potability. Water will be measured as “Potable” (1) and “Not potable” (0).  The main analytic method we intend to use will be predictive analysis to determine which factors (i.e., Chloramine, Sulfate, and Trihalomethane levels) play the most important roles in influencing water potability.  Specifically, LDA and QDA techniques will be used.

*Create 2-3 exploratory that investigate if the input variable affects the output variable. You will need to select the correct type of graph. State if based on the graph the input affects the output.*


*Implement any instance/model with the analytical tool you intend to use. This does not have to be your final model.*


```{r Loading Data}
water_data <- read.csv("water_potability.csv")
summary(water_data) #pH has 491 missing values, sulfate has 781, and Trihalomethanes had 162. 
```
```{r Data Cleaning}
water_data <- na.omit(water_data)

str(water_data)
```
##Visualizations

```{r}
plot(water_data$Chloramines,water_data$Potability)
plot(water_data$Sulfate,water_data$Potability)
plot(water_data$Conductivity,water_data$Potability)
plot(water_data$Organic_carbon,water_data$Potability)
plot(water_data$Trihalomethanes,water_data$Potability)
plot(water_data$Turbidity,water_data$Potability)
plot(water_data$ph,water_data$Potability)
```

```{r}

r <- water_data %>% 
  dplyr::select(where(is.numeric)) %>% 
  cor()

#Create the correlation heatmap using corrplot
corrplot(r)

r[,5]

abs(r[,5]) >.59 

cor(select(water_data, c(Sulfate, Solids, Hardness, ph, Chloramines, Conductivity, Organic_carbon, Turbidity, Trihalomethanes) ) )
```
##Modeling 
```{r}
model1 <- glm(Potability ~ Sulfate + Solids + Hardness + ph + Chloramines + Conductivity + Organic_carbon + Turbidity + Trihalomethanes, data = water_data)
summary(model1)
```
* Look at correlation values
* Reduce variables 
* Make visualizations
```{r}
model2 <- glm(Potability ~ Sulfate + Solids + ph + Chloramines + Conductivity + Organic_carbon + Turbidity + Trihalomethanes, data = water_data)
summary(model2)
```
```{r}
model3 <- glm(Potability ~ Solids + ph + Chloramines + Conductivity + Organic_carbon + Turbidity + Trihalomethanes, data = water_data)
summary(model3)
```
```{r}
model4 <- glm(Potability ~ Solids + ph + Chloramines + Conductivity + Organic_carbon + Turbidity, data = water_data)
summary(model4)
```
```{r}
model5 <- glm(Potability ~ Solids + ph + Chloramines + Turbidity, data = water_data)
summary(model5)

model6 <- glm(Potability ~ Solids + Chloramines + Turbidity, data = water_data)
summary(model6)

model7 <- glm(Potability ~ Solids + ph + Solids*ph + Chloramines + Turbidity, data = water_data)
summary(model7)
```
```{r Final Model}
(finalmodel <- glm(Potability ~ Solids, data = water_data))
summary(finalmodel)
```

```{r}
full.model = lm(Potability ~ Solids, data = water_data) #Nomally you need to use the lm() command
summary(full.model)

shapiro.test(full.model$residuals) #p-value < 2.2e-16

```

```{r}
Potability.null <- as.formula("Potability ~ 1")
Potability.null
```
```{r}
backward.model <-step(full.model, scope = Potability.null, direction = "backward", trace = 1)
```


```{r}
summary(backward.model)
```
```{r}
water_data <- water_data %>%
  mutate(Potability = case_when(
    Potability == 0 ~ 1, 
    Potability == 1 ~ 2
  ))
(boxcox_result <- boxcox(Potability ~ Solids, data = water_data))
```

```{r}
#SPLIT DATA INTO TRAINING AND TESTING SET
water_data$Potability <- as.factor(water_data$Potability)

str(water_data)

set.seed(1)
(sample <- sample(c(TRUE, FALSE), nrow(water_data), replace=TRUE, prob=c(0.7,0.3)))
train <- water_data[sample, ]
test <- water_data[!sample, ] 

#FIT LDA MODEL
lda_model <- lda(Potability ~ ., data = train)
lda_model

lda_predictions <- predict(lda_model, test)

lda_predicted_classes <- lda_predictions$class


qda_model <- qda(Potability ~ ., data = train)
qda_model

qda_predictions <- predict(qda_model, test)
qda_predicted_classes <- qda_predictions$class


# Load caret package for confusionMatrix
library(caret)

# Confusion matrix for LDA predictions
lda_cm <- confusionMatrix(lda_predicted_classes, test$Potability)
print(lda_cm)

# Confusion matrix for QDA predictions
qda_cm <- confusionMatrix(qda_predicted_classes, test$Potability)
print(qda_cm)


```

#59% if the observations in the training dataset were not potable and 41% were potable 
The mean amount of solids for not potable water is 21640.75 and its 22534.06 for potable water

LD1 = (0.0001148746 * Solids)

LD1 is positively correlated with Solids which means that higher levels of Solids are associated with a higher likelihood of the water being potable (1) though the effect is quite small. 















